Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-03-25 SIR_manuscript.tex
corrected A in Delta expression, white color threshold
rewrite tex in theta, change the order of legend in plot

The classical SIR model is based on the following implicit assumptions; well-mixed population, homogeneity of the population (i.e., all individuals are equaly susciptable and equaly infectious for the same length of time when infected), exponentially distributed duration of infection and large population size (see, e.g., \cite{keeling2011modeling}). In addition to these standard assumptions, our model, \ref{model}, assumes: (i) there is a single force of infection (new cases per unit time), $\Lambda$, defined as follows
\begin{equation}
\label{Lambda}
\Lambda=\beta \frac{(I\_u+\eta\_w I\_n+\eta\_w I\_p+ \eta\_c I\_c)}{N_0},
\end{equation}
across all susceptible pools with transmission rate $\beta$ and relative probability of transmission for ``waiting'' and \emph{confirmed positive} individuals, $\eta\_w$ and $\eta\_c$ respectively, (see Table~\ref{tab:params} for further details), (ii) $\eta\_c \leq \eta\_w$, i.e., the individuals awaiting test results have a higher transmission probability than the reported individuals. Note that for convential purposes in the interpretation of the results with respect to these isolation parameters, we define the transformed isolation parameters as $\theta\_w=1-\eta\_w$ and $\theta\_c=1-\eta\_c$. Thus, for instance when the awaiting people follow the isolation perfectly, $\theta\_w$ is closer to 1, while when they less follow the isolation, $\theta\_w$ is closer to 0. For this analysis, we also assume a perfectly specific test ($p_S=0$). This last assumption combined with the assumption that no individual is in waiting-for-\emph{positive} or \emph{confirmed positive} compartments, i.e., $S\_p(0)=S\_c(0)=0$, reduces the model to 10 equations (equations c and d  in \eqref{model} are eliminated).

\begin{table}[htp]
\centering
\begin{tabular}{|c|L{2in}cc|} \hline
  Symbol & Description & Unit & Value \\ \hline
  $N_0$     & Total population size & people & $10^6$ \\ \hline
  $\omega$  & Rate of onward flow from ``waiting'' to ``confirmed'' or ``untested'' compartments  & 1/day & - \\ \hline
  $\gamma$ & Recovery rate & 1/day & 1/3 \\ \hline 
  $\rho$   & Per capita testing intensity & 1/day & 0.01 \\ \hline 
  $\eta\_w$ ($\theta\_w=1-\eta\_w$)  & Relative probability of transmission for ``waiting'' individuals & - & - \\ \hline
  $\eta\_c$ ($\theta\_c=1-\eta\_c$) & Relative probability of transmission for ``confirmed positive'' individuals & - & -  \\ \hline
  $\beta$ & Transmission rate & 1/day & 0.39 \\ \hline
  $\Lambda$ & Force of infection & 1/day & - \\ \hline
  $p_S$ & Probability of positive tests for $S$ ($= 1-\textrm{specificity}$) & - & 0 \\ \hline
  $p_I$ & Probability of positive tests for $I$ ($= \textrm{sensitivity}$) & - & 1 \\ \hline
  $p_R$ & Probability of positive tests for $R$ ($= 1-\textrm{specificity}$) & - & 0.5 \\ \hline
  $w_S, w_I, w_R$ & Relative testing weights & - &
  \begin{minipage}[t]{0.21\columnwidth}%
    Random: $\{1,1,1\}$ \\ Targeted: $\{0.3,1,1\}$
  \end{minipage} \\
  \hline
  \end{tabular}
\caption{\label{tab:params} Parameters of the model \eqref{model}.\todo{figure out the best way to put the new definition $\theta=1-\eta$} in the table.}
\end{table}

\todo{the algebra needs to be changed based on JD's comment} The Disease-Free Equilibrium (DFE) for the SIR model, Eqs. \eqref{model}, is given by setting the differential equations to 0 and solving for the unknowns (\ali{is this clear or the "unkowns" needs to be defined? also * in sup for I and R?}. The DFE is
\begin{equation}
\label{dfe}
S\_n^*= \frac{\rho}{\omega} N_0, \ S\_u^*= N_0-S\_n^*, \text{~and~} I_j=R_j=0 \ \text{for all }j.
\end{equation}

The basic reproduction number, $\Rnum$, was calculated by using the next generation matrix method developed by \cite{van2002reproduction}. $\Rnum$ is
\begin{equation}
\label{R0}
\Rnum= \frac{\beta}{\gamma} (1-\Delta), 
\end{equation}
where
\begin{equation}
\begin{aligned}
\label{eq:abc}
\Delta=& 1-(A~ S\_u^* + B~ S\_n^*)~ C\\
A=& \gamma(\omega+\gamma) + (\gamma \eta\_w + \omega \eta\_c p_I) F_I, \\
B=& \big(\omega+(F_I+\gamma)\eta\_w\big) \gamma+\frac{(\eta\_w \gamma+ \eta\_c\omega) \omega p_I F_I }{\omega+\gamma}, \\ 
C=& \frac{1}{N_0 (\gamma(\omega+\gamma)+F_I(\gamma+\omega p_I))}.
\end{aligned}
\end{equation}
Further details of derivation of $\Rnum$ are provided in the appendix section.

@@

The explicit formula for the basic reproduction number, $\Rnum$ \eqref{R0}, provides an opportunity to study the influence of changes in the underlying parameters on the critical index of epidemic dynamics. We are interested in understanding the effect of parameters that can be realistically controlled by changing in isolation, \percap testing intensity and test resulting, i.e., $\eta\_c$ and $\eta\_w$, $\rho$ and $\omega$, respectively. The following Proposition is the direct result of taking partial derivative of $\Rnum$ \eqref{R0} with respect to selected parameters. Note that we used Taylor approximation of $\Rnum$ at $\rho=0$ due to the complexity of the expression and thus inconclusive.\david{in that case it would be good to establish numerically that the relationships given in the proposition are valid more generally, at least for reasonable parameter ranges}\ali{will do the simulation for testing intensity $>>1$}
See the appendix for details.

\david{More importantly, you really need to state in words what this proposition is saying.  The reader has to check the param table and think it through themselves.}
\begin{proposition}
\label{prop1}
For model \eqref{model}, and by using the expression of the basic reproduction number $\Rnum$ \eqref{R0} ,
\begin{enumerate}
\item \label{p1:eta} increasing the relative probabilities of transmission for tested individuals, increases $\Rnum$;
$$\partial{\Rnum}/\partial{\eta\_c} \geq 0 \text{~and~} \partial{\Rnum}/\partial{\eta\_w} \geq 0.$$ 
\item \label{p1:rho} higher rate of testing reduces $\Rnum$;
$$\partial{\Rnum}/\partial{\rho} \leq 0 \text{~when~} \rho \approx 0.$$
\item \label{p1:omega} The rate of test return may result in the disease becomes extinct or not;
$$\partial{\Rnum}/\partial{\omega} \text{~can be positive or negative when~} \rho \approx 0.$$
\end{enumerate}
\end{proposition}

Given that the perfect isolation occurs when no one transmits while waiting for test results (i.e., $\eta\_c=\eta\_w = 0$), Prop.~\ref{p1:eta} means that lifting the isolation from awaiting group results in an increased $\Rnum$ and consequently a greater number of infected individuals. Prop.~\ref{p1:rho} indicates that increasing the \percap testing intensity, $\rho$, reduces $\Rnum$. Lastly, Prop.~\ref{p1:omega} means that speeding up the test reporting may or may not reduce $\Rnum$, it will depend on other parameters. It is straightforward to derive \ref{p1:rho} and \ref{p1:omega} from the Taylor approximation of $\Rnum$ at $\rho=0$.
\david{can we gain some insight from the expressions for the partial derivatives in the proposition, as opposed to just their signs?}\ali{help}

\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random.pdf}
\caption{}\label{p.a}
\end{subfigure}
%
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI.pdf}
\caption{}\label{p.b}
\end{subfigure}
\caption{
{\bf A comparison of the behaviour of the basic reproduction number, $\Rnum$, between random versus targeted testing strategies at different levels of testing and isolation.}
We numerically evaluate $\Delta$ \eqref{R0}, reflecting the reduction of $\Rnum$ with respect to testing and isolation. We use the following parameters (listed in Table \ref{tab:params}):
$N_0=1 \times 10^6$, $\omega \in [0.1,2]$ 1/day, $1/\gamma= 3$ days, $\rho \in [0,0.01]$ 1/day, $\eta\_w \in [0,1]$, $\eta\_c \in [0,1]$, $\beta=0.39$ 1/day, $p_S=0$, $p_I=1$ and $p_R=0.5$. Note we are plotting $\eta\_w'=1-\eta\_w$ and $\eta\_c'=1-\eta\_c$ in the facets to better reflect the effect of isolation (0 for no effect and 1 for full effect). Contours of $\Delta$ are plotted for two testing strategies identified by a set of relative testing weights; (a) random testing where $w_S=w_I=w_R=1$ and (b) targeted testing where $w_S=0.3$ and $w_I=w_R=1$. The black solid line in each panel represents the critical contour of $\Delta=1-\frac{\gamma}{\beta}$, i.e., the $\Delta$ corresponding to the threshold of $\Rnum=1$. \bmb{we still have some collisions in the y-axis tick labels}
}
\label{pan}
\end{figure}

Our numerical solutions in support of the analytical results in Proposition \ref{prop1} are presented in Figure \ref{pan}. It can be inferred that when random testing strategy is applied, Figure \ref{p.a}, comparing to the targeted testing strategy, Figure \ref{p.b}, the critical \percap testing intensity of the corresponding panels are lower in non-random testing. Also, it appears that speeding the test reporting, i.e., increasing $\omega$, does not significantly lower $\Rnum$ when awaiting people (including people in $I\_p$ and $I\_n$ compartments) follow the isolation perfectly, i.e., $\eta\_w$ is closer to 0. However, speeding the test reporting reduces the epidemic more when the awaiting people less follow the isolation, i.e., $\eta\_w$ is closer to 1.

>>>>

The classical SIR model is based on the following implicit assumptions; well-mixed population, homogeneity of the population (i.e., all individuals are equaly susciptable and equaly infectious for the same length of time when infected), exponentially distributed duration of infection and large population size (see, e.g., \cite{keeling2011modeling}). In addition to these standard assumptions, our model, \ref{model}, assumes: (i) there is a single force of infection (new cases per unit time), $\Lambda$, defined as follows
\begin{equation}
\label{Lambda}
\Lambda=\frac{\beta}{N_0} \big(I\_u + (1-\theta\_w)(I\_n+I\_p) + (1-\theta\_c)I\_c \big),
\end{equation}
across all susceptible pools with transmission rate $\beta$ and isolation efficacy in reduction of the probability of transmission for ``waiting'' and \emph{confirmed positive} individuals, $\theta\_w$ and $\theta\_c$ respectively, (see Table~\ref{tab:params} for further details), (ii) $\theta\_c \geq \theta\_w$, i.e., the individuals awaiting test results have a higher transmission probability than the reported individuals. Thus, for instance when the awaiting people follow the isolation perfectly, $\theta\_w$ is closer to 1, while when they less follow the isolation, $\theta\_w$ is closer to 0. For this analysis, we also assume a perfectly specific test ($p_S=0$). This last assumption combined with the assumption that no individual is in waiting-for-\emph{positive} or \emph{confirmed positive} compartments, i.e., $S\_p(0)=S\_c(0)=0$, reduces the model to 10 equations (equations c and d  in \eqref{model} are eliminated).

\begin{table}[htp]
\centering
\begin{tabular}{|c|L{2in}cc|} \hline
  Symbol & Description & Unit & Value \\ \hline
  $N_0$     & Total population size & people & $10^6$ \\ \hline
  $\omega$  & Rate of onward flow from ``waiting'' to ``confirmed'' or ``untested'' compartments  & 1/day & - \\ \hline
  $\gamma$ & Recovery rate & 1/day & 1/3 \\ \hline 
  $\rho$   & Per capita testing intensity & 1/day & 0.01 \\ \hline 
  $\theta\_w$ & Isolation efficacy in reduction of the probability of transmission for ``waiting'' individuals & - & - \\ \hline
  $\theta\_c$ & Isolation efficacy in reduction of the probability of transmission for ``confirmed positive'' individuals & - & -  \\ \hline
  $\beta$ & Transmission rate & 1/day & 0.39 \\ \hline
  $\Lambda$ & Force of infection & 1/day & - \\ \hline
  $p_S$ & Probability of positive tests for $S$ ($= 1-\textrm{specificity}$) & - & 0 \\ \hline
  $p_I$ & Probability of positive tests for $I$ ($= \textrm{sensitivity}$) & - & 1 \\ \hline
  $p_R$ & Probability of positive tests for $R$ ($= 1-\textrm{specificity}$) & - & 0.5 \\ \hline
  $w_S, w_I, w_R$ & Relative testing weights & - &
  \begin{minipage}[t]{0.21\columnwidth}%
    Random: $\{1,1,1\}$ \\ Targeted: $\{0.3,1,1\}$
  \end{minipage} \\
  \hline
  \end{tabular}
\caption{\label{tab:params} Parameters of the model \eqref{model}.}
\end{table}

\todo{the algebra needs to be changed based on JD's comment} The Disease-Free Equilibrium (DFE) for the SIR model, Eqs. \eqref{model}, is given by setting the differential equations to 0 and solving for the unknowns (\ali{is this clear or the "unkowns" needs to be defined? also * in sup for I and R?}. The DFE is
\begin{equation}
\label{dfe}
S\_n^*= \frac{\rho}{\omega} N_0, \ S\_u^*= N_0-S\_n^*, \text{~and~} I_j=R_j=0 \ \text{for all }j.
\end{equation}
For simplification of the analysis, we adopt the following change $x\_u^* = S\_u^*/N_0$ and $x\_n^* = S\_n^*/N_0$.

The basic reproduction number, $\Rnum$, was calculated by using the next generation matrix method developed by \cite{van2002reproduction}. $\Rnum$ is
\begin{equation}
\label{R0}
\Rnum= \frac{\beta}{\gamma} (1-\Delta), 
\end{equation}
where
\begin{equation}
\begin{aligned}
\label{eq:abc}
\Delta=& (A~ \theta\_w + B~ \theta\_c)~ C\\
A=& \gamma \big( (\gamma \omega+ \omega p_I F_I+\gamma^2) x_n^* + (\omega+\gamma) F_I \big)\\
B=& \omega p_I F_I (\gamma x_u^*+\omega) \\ 
C=& \frac{1}{(\omega+\gamma) \big(\gamma(\omega+\gamma)+F_I(\gamma+\omega p_I) \big)}.
\end{aligned}
\end{equation}

Further details of derivation of $\Rnum$ are provided in the appendix section.

@@

The explicit formula for the basic reproduction number, $\Rnum$ \eqref{R0}, provides an opportunity to study the influence of changes in the underlying parameters on the critical index of epidemic dynamics. We are interested in understanding the effect of parameters that can be realistically controlled by changing in isolation, \percap testing intensity and test resulting, i.e., $\theta\_c$ and $\theta\_w$, $\rho$ and $\omega$, respectively. The following Proposition is the direct result of taking partial derivative of $\Rnum$ \eqref{R0} with respect to selected parameters. Note that we used Taylor approximation of $\Rnum$ at $\rho=0$ due to the complexity of the expression and thus inconclusive.\david{in that case it would be good to establish numerically that the relationships given in the proposition are valid more generally, at least for reasonable parameter ranges}\ali{will do the simulation for testing intensity $>>1$}
See the appendix for details.

\david{More importantly, you really need to state in words what this proposition is saying.  The reader has to check the param table and think it through themselves.}
\begin{proposition}
\label{prop1}
For model \eqref{model}, and by using the expression of the basic reproduction number $\Rnum$ \eqref{R0} ,
\begin{enumerate}
\item \label{p1:eta} increasing the relative probabilities of transmission for tested individuals, increases $\Rnum$;
$$\partial{\Rnum}/\partial{\theta\_c} \leq 0 \text{~and~} \partial{\Rnum}/\partial{\theta\_w} \leq 0.$$ 
\item \label{p1:rho} higher rate of testing reduces $\Rnum$;
$$\partial{\Rnum}/\partial{\rho} \leq 0 \text{~when~} \rho \approx 0.$$
\item \label{p1:omega} The rate of test return may result in the disease becomes extinct or not;
$$\partial{\Rnum}/\partial{\omega} \text{~can be positive or negative when~} \rho \approx 0.$$
\end{enumerate}
\end{proposition}

Given that the perfect isolation occurs when no one transmits while waiting for test results (i.e., $\theta\_c=\theta\_w = 1$), Prop.~\ref{p1:eta} means that lifting the isolation from awaiting group results in an increased $\Rnum$ and consequently a greater number of infected individuals. Prop.~\ref{p1:rho} indicates that increasing the \percap testing intensity, $\rho$, reduces $\Rnum$. Lastly, Prop.~\ref{p1:omega} means that speeding up the test reporting may or may not reduce $\Rnum$, it will depend on other parameters. It is straightforward to derive \ref{p1:rho} and \ref{p1:omega} from the Taylor approximation of $\Rnum$ at $\rho=0$.
\david{can we gain some insight from the expressions for the partial derivatives in the proposition, as opposed to just their signs?}\ali{help}

\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random.pdf}
\caption{}\label{p.a}
\end{subfigure}
%
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI.pdf}
\caption{}\label{p.b}
\end{subfigure}
\caption{
{\bf A comparison of the behaviour of the basic reproduction number, $\Rnum$, between random versus targeted testing strategies at different levels of testing and isolation.}
We numerically evaluate $\Delta$ \eqref{R0}, reflecting the reduction of $\Rnum$ with respect to testing and isolation. We use the following parameters (listed in Table \ref{tab:params}):
$N_0=1 \times 10^6$, $\omega \in [0.1,2]$ 1/day, $1/\gamma= 3$ days, $\rho \in [0,0.01]$ 1/day, $\theta\_w$ and $\theta\_c$ vary between 0 and 1 with 0 for no effect and 1 for full effect of isolation on the transmission probability, $\beta=0.39$ 1/day, $p_S=0$, $p_I=1$ and $p_R=0.5$. Contours of $\Delta$ are plotted for two testing strategies identified by a set of relative testing weights; (a) random testing where $w_S=w_I=w_R=1$ and (b) targeted testing where $w_S=0.3$ and $w_I=w_R=1$. The black solid line in each panel represents the critical contour of $\Delta=1-\frac{\gamma}{\beta}$, i.e., the $\Delta$ corresponding to the threshold of $\Rnum=1$. \bmb{we still have some collisions in the y-axis tick labels}
}
\label{pan}
\end{figure}

Our numerical solutions in support of the analytical results in Proposition \ref{prop1} are presented in Figure \ref{pan}. It can be inferred that when random testing strategy is applied, Figure \ref{p.a}, comparing to the targeted testing strategy, Figure \ref{p.b}, the critical \percap testing intensity of the corresponding panels are lower in non-random testing. Also, it appears that speeding the test reporting, i.e., increasing $\omega$, does not significantly lower $\Rnum$ when awaiting people (including people in $I\_p$ and $I\_n$ compartments) follow the isolation perfectly, i.e., $\theta\_w$ is closer to 1. However, speeding the test reporting reduces the epidemic more when the awaiting people less follow the isolation, i.e., $\theta\_w$ is closer to 0.
